---
title: "Store XML in a Database"
output: html_notebook
### Author: Zhang, Jinwei
### Course: CS5200
### Term: 2023 Spring
---

## Q1: Create a normalized (BCNF) relational schema
Assumptions about XML data:
1. The schema has following functional dependencies:
   (1) bid -> title, eid
   (2) title -> author, genre, price, description
   (3) eid -> edition, publish_date
2. Only the books with different editions can have the same title
3. Each book has only one author, one genre
4. Each book must have an edition(default 1) and publish_date
```{r}
library(knitr)
imageURL <- "https://i.imgur.com/l3ng4N5.jpg"
knitr::include_graphics(imageURL)
```


## Q2: Create a SQLite database that implements the schema
### connect db
```{r}
library(RSQLite)

# connect to sqlite db
wd = getwd()
fpath = paste0(wd, "/")
dbfile = "xml_sqlite.db"

dbcon <- dbConnect(RSQLite::SQLite(), paste0(fpath, dbfile))
```

### enable foreign key constraint
```{sql connection=dbcon}
PRAGMA foreign_keys = ON
```

### drop table if exists
```{sql connection=dbcon}
DROP TABLE IF EXISTS Book;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS BookDetail;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS EditionDetail;
```

### create tables
```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS BookDetail(
  title TEXT NOT NULL PRIMARY KEY,
  author TEXT,
  genre TEXT,
  price NUMERIC,
  description TEXT
);
```

```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS EditionDetail(
  eid INTEGER NOT NULL PRIMARY KEY,
  edition INTEGER NOT NULL,
  publish_date DATE NOT NULL,
  UNIQUE (edition, publish_date)
);
```

```{sql connection=dbcon}
CREATE TABLE IF NOT EXISTS Book(
  bid TEXT NOT NULL PRIMARY KEY,
  title TEXT NOT NULL,
  eid INTEGER NOT NULL,
  FOREIGN KEY (title) REFERENCES BookDetail(title),
  FOREIGN KEY (eid) REFERENCES EditionDetail(eid)
);
```

## Q3: Load the XML data from the file into R data frames
### load XML into DOM
```{r}
library(XML)
wd = getwd()
fpath = paste0(wd, "/")
# read XML file
xmlFile <- paste0(fpath, "Books-v3.xml")
xmlObj <- xmlParse(xmlFile)
# get the size of DOM
r <- xmlRoot(xmlObj)
size <- xmlSize(r)
```

### create internal data frames
```{r}
# stringsAsFactors=F tells R not to convert character vectors into factors when creating a data frame
BookDetail <- data.frame(title = character(),
                         author = character(), 
                         genre = character(),
                         price = numeric(),
                         description = character(),
                         stringsAsFactors = F)

EditionDetail <- data.frame(eid = integer(),
                            edition = integer(), 
                            publish_date = character(),
                            stringsAsFactors = F)

Book <- data.frame(bid = integer(),
                   title = character(), 
                   eid = integer(),
                   stringsAsFactors = F)
```

### functions to check duplicates
```{r}
checkTitle <- function(title) {
  index <- match(title, BookDetail$title)
  if (!is.na(index)) {
    return(index)
  } else {
    return(-1)
  }
}

checkEdition <- function(edition, publish_date) {
  # check if EditionDetail data frame contains the row
  row_index <- which(EditionDetail$edition == edition & EditionDetail$publish_date == publish_date)
  if (length(row_index) > 0) {
    return(row_index)
  } else {
    return(-1)
  }
}
```

### support function to parse data
```{r}
parseBookDetail <- function(curNode)
{
  # if using //book/title, even pass the single node, it will extract titles of the whole document
  title <- xpathSApply(curNode, "./title", xmlValue)
  author <- xpathSApply(curNode, "./author", xmlValue)
  genre <- xpathSApply(curNode, "./genre", xmlValue)
  price <- as.numeric(xpathSApply(curNode, "./price", xmlValue))
  description <- xpathSApply(curNode, "./description", xmlValue)
  
  newBookDetail.df <- data.frame(title, author, genre, price, description, stringsAsFactors = F)
  return(newBookDetail.df)
}
```

### iterate over nodes and fill the data frames
```{r}
bdIdx <- 1
edIdx <- 1
bkIdx <- 1
for (i in 1:size){
  # get current book
  curBookNode <- r[[i]]
  
  # new title, add to BookDetail data frame
  title <- xpathSApply(curBookNode, "./title", xmlValue)
  if(checkTitle(title) == -1){
    BookDetail[bdIdx,] <- parseBookDetail(curBookNode)
    bdIdx <- bdIdx + 1
  }
  
  edition <- xpathSApply(curBookNode, "./edition", xmlValue)
  publish_date <- xpathSApply(curBookNode, "./publish_date", xmlValue)
  # no edition, default 1
  if(length(edition) != 1){
    edition <- as.integer(1)
  } else {
    edition <- as.integer(edition)
  }
  # new edition detail, add to EditionDetail data frame
  if(checkEdition(edition, publish_date) == -1){
    eid <- as.integer(edIdx)
    EditionDetail[edIdx,] <- data.frame(eid, edition, publish_date, stringsAsFactors = F)
    edIdx <- edIdx + 1
  }
  
  # add to Book dataframe
  bid <- xmlAttrs(r[[i]])[[1]]
  # get the eid
  eid <- checkEdition(edition, publish_date)
  Book[bkIdx,] <- data.frame(bid, title, eid, stringsAsFactors = F)
  bkIdx <- bkIdx + 1
}
```


### display data frames
```{r}
Book
BookDetail
EditionDetail
```

## Q4: write data from dataframes to SQLite db
```{r}
dbWriteTable(dbcon, "BookDetail", BookDetail, overwrite = T)
dbWriteTable(dbcon, "EditionDetail", EditionDetail, overwrite = T)
dbWriteTable(dbcon, "Book", Book, overwrite = T)
```

### show tables
```{sql connection=dbcon}
SELECT * FROM Book;
```

```{sql connection=dbcon}
SELECT * FROM BookDetail;
```

```{sql connection=dbcon}
SELECT * FROM EditionDetail;
```

## Q5: write SQL queries
A: What is the number of genres have at least three books?
```{sql connection=dbcon}
SELECT COUNT(genre) AS num FROM 
  (SELECT genre FROM Book b
    JOIN BookDetail bd ON b.title = bd.title
    GROUP BY genre
    HAVING COUNT(bid) >= 3)
```
B: What is the oldest year in which a publication was published?
```{sql connection=dbcon}
SELECT MIN(year) AS oldest_year FROM
  (SELECT SUBSTR(publish_date, 1, 4) AS year FROM EditionDetail);
```
C: Find the number of books and average price for each genre.
```{sql connection=dbcon}
SELECT genre, COUNT(bid) AS total, ROUND(AVG(price), 2) AS avg_price FROM Book b
  JOIN BookDetail bd ON b.title = bd.title
  GROUP BY genre;
```
D: List the title and author of all books that are less than 0.8 * AVG or more than 1.2 * AVG, where AVG is the average price of all books. 
```{sql connection=dbcon}
SELECT DISTINCT title, author FROM BookDetail
  WHERE price < 0.8*(SELECT AVG(price) FROM Book b 
                      JOIN BookDetail bd ON b.title = bd.title) 
  OR price > 1.2*(SELECT AVG(price) FROM Book b 
                    JOIN BookDetail bd ON b.title = bd.title);
```

### discconect db
```{r}
dbDisconnect(dbcon)
```


