---
title: "Query XML"
output: html_notebook
### Author: Zhang, Jinwei
### Course: CS5200
### Term: 2023 Spring
---

### read XML file
```{r}
library(XML)

# get the file path
wd = getwd()
fpath = paste0(wd, "/")
xmlFile <- paste0(fpath, "Books-v3.xml")

# read XML file
xmlObj <- xmlParse(xmlFile)
xmlObj
```

## Query 1
What are the titles of all books written by "Galos, Mike"? List the titles and the prices.
```{r}
xpathEx <- "//book[author='Galos, Mike']/title/text() | //book[author='Galos, Mike']/price/text()"
res <- xpathSApply(xmlObj, xpathEx, xmlValue)
res
```

## Query 2
In what genre is the book with id "bk109".
```{r}
xpathEx <- "//book[@id='bk109']/genre/text()"
res <- xpathSApply(xmlObj, xpathEx, xmlValue)
res
```

## Query 3
What are the titles of the books that sell for the lowest price? (Use the min aggregation function)
```{r}
xpathEx <- "//book[price=min(//book/price)]/title/text()"
res <- xpathSApply(xmlObj, xpathEx, xmlValue)
res
```

## Query 4
List the title and prices of all books in the "Computer" genre.
```{r}
xpathEx <- "//book[genre='Computer']/title/text() | //book[genre='Computer']/price/text()"
res <- xpathSApply(xmlObj, xpathEx, xmlValue)
res
```

## Query 5
What is the number of all books that cost less than $5? (Use the count aggregation function)
```{r}
xpathEx <- "count(//book[price < 5])"
res <- xpathSApply(xmlObj, xpathEx, xmlValue)
res
```
