---
title: "R Notebook"
output: html_notebook
### Author: Zhang, Jinwei
### Email: zhang.jinw@northeastern.edu
### Course: CS5200
### Term: 2023 Spring
---

# create connection to MySQL db
```{r}
library(DBI)
library(RMySQL)
# use readr to load csv file
library(readr)
# forces sqldf to use SQLite as its backing store
library(sqldf)
options(sqldf.driver = 'SQLite')

dbcon <- dbConnect(MySQL(), user = 'root', password = 'root',
                 dbname = 'bird_strike', host = 'localhost', port = 3306)
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS incidents;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS airports;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS conditions;
```

# Task1B: Create airports table
# The assumption is that using only airport_name can't determine the whole record
# airports in different state may have same name
```{sql connection=dbcon}
CREATE TABLE airports(
  aid INT PRIMARY KEY AUTO_INCREMENT,
  airport_name VARCHAR(50),
  airport_code VARCHAR(30) DEFAULT '',
  state VARCHAR(30)
);
```

# Task1D: Create conditions lookup table
# NOTE: 'condition' is the key word
```{sql connection=dbcon}
CREATE TABLE conditions(
  cid INT PRIMARY KEY AUTO_INCREMENT,
  sky_conditions VARCHAR(30) UNIQUE,
  explanation VARCHAR(50) DEFAULT ''
);
```

# Task1A,C: Create incidents table
```{sql connection=dbcon}
CREATE TABLE incidents(
  rid INT PRIMARY KEY,
  flight_date DATE,
  origin_id INT,
  airline VARCHAR(50),
  aircraft VARCHAR(30),
  flight_phase VARCHAR(30),
  altitude INT,
  condition_id INT,
  warning BOOLEAN,
  FOREIGN KEY(origin_id) REFERENCES airports(aid),
  FOREIGN KEY(condition_id) REFERENCES conditions(cid)
);
```

# --------------------------
# Task2: populate tables
# load csv data
```{r}
bird_strikes <- read.csv("BirdStrikesData-V2.csv")
# remove all military flights(89 rows) -> 25469 rows
bird_strikes <- bird_strikes[!bird_strikes$airline == 'MILITARY',]
# remove records without flight information(aircraft,flight_date...) -> 25340 rows(129 rows removed)
# NOTE: '&' or '|' will evaluate the whole vector, while '&&' and '||' will only evaluate the first element
bird_strikes <- bird_strikes[!(bird_strikes$aircraft == '' & bird_strikes$flight_date == ''),]
bird_strikes 
```

# build airports data frame
```{r}
airports <- sqldf('SELECT DISTINCT airport AS airport_name, origin AS state 
                    FROM bird_strikes
                    ORDER BY state')

# omit rows with airport_name be empty
# airports <- airports[!(airports$airport_name == ''),]

# if the state is N/A, the airport is not reside within USA, leave it empty
airports$state[airports$state == 'N/A'] <- ''
airports
```

# build conditions data frame
```{r}
conditions <- sqldf('SELECT DISTINCT sky_conditions FROM bird_strikes')
conditions
```

# populate airports table
```{r}
populate_airports <- function() {
  # iterate airports data frame
  for (i in 1:nrow(airports)) {
    # use "" to include value, value may contain ' \' '
    airport_name <- paste0('"', airports[i,1], '"')
    state <- paste0('"', airports[i,2], '"')
    cmd<-paste0('INSERT INTO airports(airport_name, state) VALUES (', airport_name, ',' , state, ')')
    dbExecute(dbcon,cmd)
  }
}
  
populate_airports()
```

# populate conditions table
```{r}
populate_conditions <- function() {
  # iterate conditions data frame
  for (i in 1:nrow(conditions)) {
    # use "" to include value
    sky_conditions <- paste0('"', conditions[i,1], '"')
    cmd<-paste0('INSERT INTO conditions(sky_conditions) VALUES (', sky_conditions, ')')
    dbExecute(dbcon,cmd)
  }
}
  
populate_conditions()
```

# populate incidents table
# this may take half a minute
```{r}
populate_incidents <- function() {
  # iterate incidents data frame
  # for (i in 1:10)
  for (i in 1:nrow(bird_strikes)){
    # rid
    rid <- bird_strikes[i, 1]
    # rid <- paste0('"', bird_strikes[i, 1], '"')
    
    # modify flight_date
    tmp_date <- strsplit(bird_strikes[i, 7], ' ')[[1]] # omid time
    tmp_date <- strsplit(tmp_date[[1]], '/')[[1]] # split y,m,d
    # MySQL permits a “relaxed” format for values specified as strings
    # for example, any punctuation chracter may be used as the delimiter;
    # moreover, it will autofill prefix 0 for month and day(1-9)
    flight_date <- paste(tmp_date[[3]], tmp_date[[1]], tmp_date[[2]], sep = '-')
    # flight_date <- paste(paste0('20', tmp_date[[3]]), tmp_date[[1]], tmp_date[[2]], sep = '-')
    flight_date <- paste0('"', flight_date, '"')
    
    # query airports table to get origin_id(aid), 
    # both the airport_name and state name should be matched
    airport_name <- bird_strikes[i, 3]
    state <- bird_strikes[i, 10]
    # modify state with 'N/A' value to empty string
    if(state == 'N/A'){
      state <- ''
    }
    aid_query<-paste0('SELECT aid FROM airports WHERE airport_name =', '"', airport_name, '"', 'AND state =', '"', state, '"')
    aid <- dbGetQuery(dbcon, aid_query);
    origin_id <- aid[1, 1]
    
    # airline
    airline <- paste0('"', bird_strikes[i, 9], '"')
    
    # aircraft
    aircraft <- paste0('"', bird_strikes[i, 2], '"')
    
    # modify flight_phase
    # Assumption: 
    # 1. 'Take-off run' belongs to takeoff phase
    # 2. 'Climb', 'Descent', 'Approach' belongs to inflight phase
    # 3. 'Landing Roll' belongs to landing phase
    # 4. 'Taxi' can be Taxi-out and Taxi-in; 'Parked' is at the very end of flight phase
    #    thus they belongs to unknown phase
    flight_phase <- bird_strikes[i, 11]
    if(flight_phase == 'Take-off run'){
      flight_phase <- 'takeoff'
    } else if (flight_phase == 'Climb' | flight_phase == 'Descent' | flight_phase == 'Approach'){
      flight_phase <- 'inflight'
    } else if (flight_phase == 'Landing Roll'){
      flight_phase <- 'landing'
    } else {
      # Taxi && Parked
      flight_phase <- 'unknown'
    }
    flight_phase <- paste0('"', flight_phase, '"')
    
    # altitude, no need to include it with ""
    altitude_str <- bird_strikes[i, 18]
    # get the first splitted subset
    splitted_str <- strsplit(altitude_str, ',')[[1]] 
    altitude <- ''
    # convert '1,500' into '1500'
    for(str in splitted_str){
      altitude <- paste0(altitude, str)
    }
    
    # query conditions table to get condition_id(cid)
    cid_query<-paste0('SELECT cid FROM conditions WHERE sky_conditions=', '"', bird_strikes[i, 15], '"')
    cid <- dbGetQuery(dbcon, cid_query);
    condition_id <- cid[1, 1]
    
    # warning, no need to include it with ""
    warning <- bird_strikes[i, 17]
    if(warning == 'N'){
      warning <- 'FALSE'
    } else {
      warning <- 'TRUE'
    }
    
    # execute statement
    cmd<-paste0('INSERT INTO incidents VALUES (', rid, ',', flight_date, ',', origin_id, ',', airline, ',', aircraft, ',',
                flight_phase, ',', altitude, ',', condition_id, ',', warning, ')')
    dbExecute(dbcon,cmd)
  }
}
  
populate_incidents()
```

# ---------------------------------------------------------------------------------
# Task3: Show that the loading of the data worked by displaying parts of each table 
# (do not show the entire tables).

# airports table
# leave airport_code attribute empty for future expansion
```{sql connection=dbcon}
SELECT * FROM airports LIMIT 50;
```

# conditions table(only 3 records, show the entire table)
# leave the explanation attribute empty for future expansion
```{sql connection=dbcon}
SELECT * FROM conditions;
```

# incidents table
```{sql connection=dbcon}
SELECT * FROM incidents LIMIT 50;
```

# display joint tables
```{sql connection=dbcon}
SELECT * FROM incidents i
  JOIN airports a ON i.origin_id = a.aid
  JOIN conditions c ON c.cid = i.condition_id
  ORDER BY rid
  LIMIT 50;
```

# -----------------------------------------------------------------
# Task4: find the top-10 airlines with the most number of incidents.
```{sql connection=dbcon}
SELECT airline, COUNT(rid) AS total FROM incidents 
  GROUP BY airline
  ORDER BY total DESC
  LIMIT 10;
```

# -----------------------------------------------------------------------------------
# Task5: find the flight phase that had an above average number bird strike incidents
# (during any flight phase).
# NOTE: MySQL requires the derived table to have its own alias
```{sql connection=dbcon}
SELECT flight_phase, COUNT(rid) AS total FROM incidents AS i
  GROUP BY flight_phase
  HAVING total > (SELECT AVG(cnt) 
                    FROM (SELECT COUNT(rid) AS cnt 
                            FROM incidents
                            GROUP BY flight_phase) AS sub);
```

# only for testing, not the answer of Task5
```{sql connection=dbcon}
SELECT flight_phase, COUNT(rid) AS cnt FROM incidents 
    GROUP BY flight_phase;
```

```{sql connection=dbcon}
SELECT AVG(cnt) AS average FROM (SELECT COUNT(rid) AS cnt 
                                    FROM incidents 
                                    GROUP BY flight_phase) AS sub;
```

# -----------------------------------------------------------------------------------
# Task6:find the maximum number of bird strike incidents by month (across all years). 
# Include all airlines and all flights.
# This query can help answer the question which month, historically, is the most dangerous for bird strikes.
```{sql connection=dbcon}
SELECT SUBSTR(flight_date, 6, 2) AS month, COUNT(rid) AS total FROM incidents
  GROUP BY month
  HAVING total = (SELECT MAX(cnt) 
                    FROM (SELECT COUNT(rid) AS cnt FROM incidents
                            GROUP BY SUBSTR(flight_date, 6, 2)) AS sub);
```

# only for testing, not the answer of Task6
```{sql connection=dbcon}
SELECT SUBSTR(flight_date, 6, 2) AS month, COUNT(rid) AS total FROM incidents
  GROUP BY month
  ORDER BY month;
```

```{sql connection=dbcon}
SELECT MAX(cnt) FROM (SELECT COUNT(rid) AS cnt FROM incidents
                        GROUP BY SUBSTR(flight_date, 6, 2)) AS sub;
```

# ------------------------------------------------------------------------------------------------------------
# Task7: Build a column chart that visualizes the number of bird strikes incidents per year from 2005 to 2011. 
# Adorn the graph with appropriate axis labels, titles, legend, data labels, etc.
```{r}
# import ggplot2 library for drawing column graph
library(ggplot2)
cmd <- "SELECT SUBSTR(flight_date, 1, 4) AS year, COUNT(rid) as total FROM incidents 
          GROUP BY year
          HAVING year >= '2005' AND year <= '2011'
          ORDER BY year;"
rs <- dbGetQuery(dbcon, cmd)
rs

# set the stat = 'identity' tells ggplot2 that the y-value is provided
# instead of counting the # of rows for x-value
# add column bar, data labels
graph <- ggplot(rs, aes(x = year, y = total)) + geom_bar(stat='identity', width = 0.5) + geom_text(aes(label = total), vjust = -0.5)
# add axis labels, titles
graph <- graph + labs(title = 'Number of Bird Strikes Incidents from 2005 to 2011', 
                      x = 'Year', y = 'Total Number of Incidents') 
graph
```

# only for testing, not the answer of Task7
```{sql connection=dbcon}
SELECT SUBSTR(flight_date, 1, 4) AS year, COUNT(rid) as total FROM incidents 
  GROUP BY year
  HAVING year >= '2005' AND year <= '2011'
  ORDER BY year;
```

# -----------------------------------------------------------------------------------
# Task8: Create a stored procedure in MySQL that adds a new incident to the database. 
# You may decide what you need to pass to the stored procedure to add a bird strike incident 
# and you must account for there being potentially a new airport. 
# After insertion, show (in R) that your procedure worked.
```{sql connection=dbcon}
DROP PROCEDURE IF EXISTS insert_incident_data;
```

# create stored procedure
```{sql connection=dbcon}
CREATE PROCEDURE insert_incident_data(
  IN rid INT,
  IN flight_date DATE,
  IN origin_id INT,
  IN airline VARCHAR(50),
  IN aircraft VARCHAR(30),
  IN flight_phase VARCHAR(30),
  IN altitude INT,
  IN condition_id INT,
  IN warning BOOLEAN
)
BEGIN
	DECLARE originCnt INT;
	DECLARE conditionCnt INT;
  
	SELECT COUNT(aid) INTO originCnt
	  FROM airports
	  WHERE aid = origin_id;

  SELECT COUNT(cid) INTO conditionCnt
	  FROM conditions
	  WHERE cid = condition_id;
	  
	-- check if origin_id and condition_id exists
	IF originCnt != 1 THEN 
	  -- '45000' unhandled user-defined exception.
		SIGNAL SQLSTATE '45000'
			SET MESSAGE_TEXT = 'Origin id not found in airports table';
	ELSEIF conditionCnt != 1 THEN
    SIGNAL SQLSTATE '45000'
			SET MESSAGE_TEXT = 'Condition id not found in conditions table';
  ELSE
    -- no conflicts, insert data into incidents table
    INSERT INTO incidents 
      VALUES (rid, flight_date, origin_id, airline, aircraft, flight_phase, altitude, condition_id, warning);
	END IF;
	
END

```

# insert data with new(non-exist) origin_id
# Error 'Origin id not found in airports table' will appear
```{sql connection=dbcon}
CALL insert_incident_data(1, '2000-06-18', 10000, 'test airline1', 'test aircraft1', 'inflight', 1500, 2, TRUE);
```

# insert data with new(non-exist) condition_id
# Error 'Condition id not found in conditions table' will appear
```{sql connection=dbcon}
CALL insert_incident_data(2, '2005-11-01', 765, 'test airline2', 'test aircraft2', 'landing', 0, 5, FALSE);
```

# insert data with valid values
```{sql connection=dbcon}
CALL insert_incident_data(3, '2005-11-01', 765, 'test airline3', 'test aircraft3', 'takeoff', 0, 1, FALSE);
```

# show new inserted data 
```{sql connection=dbcon}
SELECT * FROM incidents i
  JOIN airports a ON i.origin_id = a.aid
  JOIN conditions c ON i.condition_id = c.cid
  WHERE rid = 3;
```

# disconnect
```{r}
dbDisconnect(dbcon)
```