---
title: "Query a Database with SQL"
output: html_notebook
### Author: Zhang, Jinwei
### Course: CS5200
### Term: 2023 Spring
---

# create connection to db
```{r}
library(RSQLite)

# get current work directory
wd = getwd()
fpath = paste0(wd, "/")
dbfile = "MediaDB.db"

# create db or get db connection
dbcon <- dbConnect(RSQLite::SQLite(), paste0(fpath, dbfile))
```

```{sql connection=dbcon}
SELECT * FROM customers;
```

```{sql connection=dbcon}
SELECT * FROM albums;
```

```{sql connection=dbcon}
SELECT * FROM artists;
```

```{sql connection=dbcon}
SELECT * FROM tracks;
```

```{sql connection=dbcon}
SELECT * FROM media_types;
```

```{sql connection=dbcon}
SELECT * FROM genres;
```

```{sql connection=dbcon}
SELECT * FROM employees;
```

# Q1.What are the last names, city, and state of all customer who made at least one purchase?
```{sql connection=dbcon}
SELECT LastName, City, State FROM customers
  WHERE CustomerId IN(SELECT DISTINCT CustomerId FROM invoices);
```

# Q2.What are the titles of each albums and the number of tracks on each album?
```{sql connection=dbcon}
SELECT a.Title, COUNT(t.TrackId) AS TotalNum FROM albums AS a
  INNER JOIN tracks AS t ON a.AlbumId = t.AlbumId
  GROUP BY a.AlbumId;
```

# Q3.List the number of tracks by media type.
```{sql connection=dbcon}
SELECT m.MediaTypeId, m.Name, COUNT(t.TrackId) AS TrackNum FROM media_types AS m
  INNER JOIN tracks AS t ON m.MediaTypeId = t.MediaTypeId
  GROUP BY m.MediaTypeId;
```

# Q4.What is the average age of the employees? 
# Hint: SQLite does not support DATEDIFF but does support STRFTIME
# Compare year, then compare month-day to calculate the offset
```{sql connection=dbcon}
SELECT AVG((strftime('%Y', 'now') - strftime('%Y', BirthDate)) 
       - (strftime('%m-%d', 'now') < strftime('%m-%d', BirthDate))) AS AvgAge
  FROM employees;
```

# NOTE: only for testing age offset, not the answer of Q4
```{sql connection=dbcon}
SELECT EmployeeId, BirthDate, 
  (strftime('%m-%d', 'now') < strftime('%m-%d', BirthDate)) AS Diff
  FROM employees;
```

# Q5.What are the total number of unique customers for each US state (e.g., MA, CT, etc.), 
# ordered alphabetically by state, who made at least one purchase 
# (i.e., do not consider customers who never made a purchase)?
```{sql connection=dbcon}
SELECT c.State, COUNT(DISTINCT c.CustomerId) AS total FROM customers AS c
  INNER JOIN invoices AS i ON c.CustomerId = i.CustomerId
  GROUP BY c.State
  HAVING c.State IS NOT NULL
  ORDER BY c.State;
```

# NOTE: only for testing, not the answer of Q5
```{sql connection=dbcon}
SELECT DISTINCT c.CustomerId, c.State FROM customers AS c
  INNER JOIN invoices AS i ON c.CustomerId = i.CustomerId
  ORDER BY c.State;
```

# Q6.How many customers never made a purchase, 
# i.e., they appear in the customers table but not in the invoices table.
```{sql connection=dbcon}
SELECT COUNT(CustomerId) AS total FROM customers
  WHERE CustomerId NOT IN(SELECT DISTINCT CustomerId FROM invoices);
```

# NOTE: only for testing, not the answer of Q6
```{sql connection=dbcon}
SELECT DISTINCT i.CustomerId FROM invoices AS i;
```

# Q7.How many albums contain the substring "symphony" in the album title?
# LIKE in Sqlite is case-insensitive by default
# but can use lower() to convert title to lowercase
```{sql connection=dbcon}
SELECT COUNT(Title) AS total FROM albums
  WHERE LOWER(Title) LIKE '%symphony%';
```

# Q8.What is the total "time" of all tracks per artist, 
# i.e., add up the "Milliseconds" column in tracks? 
# Display the artist's name and the time in hours rounded to two decimals.
# artists with no tracks should also be displayed
```{sql connection=dbcon}
SELECT art.Name, ROUND(SUM(t.Milliseconds) * 2.778E-7, 2) AS Hours
  FROM tracks AS t
  INNER JOIN albums AS alb ON t.AlbumId = alb.AlbumId
  INNER JOIN artists AS art ON alb.ArtistId = art.ArtistId
  GROUP BY art.ArtistId
  UNION
  SELECT Name, 0.00 AS Hours FROM artists
  WHERE ArtistId NOT IN(SELECT DISTINCT ArtistId FROM albums);
```

# Q8.What is the total "time" of all tracks per artist, 
# i.e., add up the "Milliseconds" column in tracks? 
# Display the artist's name and the time in hours rounded to two decimals.
# artists with no tracks should also be displayed

# use right join
```{sql connection=dbcon}
SELECT art.Name, ROUND(SUM(IFNULL(t.Milliseconds, 0)) * 2.778E-7, 2) AS Hours
  FROM tracks AS t
  INNER JOIN albums AS alb ON t.AlbumId = alb.AlbumId
  RIGHT JOIN artists AS art ON alb.ArtistId = art.ArtistId
  GROUP BY art.ArtistId
  ORDER BY art.Name;
```

# Q9.Which artist(s) published the most tracks, 
# regardless of type and format? List the artist(s) name(s).
```{sql connection=dbcon}
SELECT art.Name FROM tracks AS t
  INNER JOIN albums AS alb ON t.AlbumId = alb.AlbumId
  INNER JOIN artists AS art ON alb.ArtistId = art.ArtistId
  GROUP BY art.ArtistId
  HAVING COUNT(t.TrackId) = (SELECT MAX(total) 
                              FROM (SELECT COUNT(t.trackId) AS total 
                                      FROM tracks AS t
                                      INNER JOIN albums AS alb ON t.AlbumId = alb.AlbumId
                                      GROUP BY alb.ArtistId));
```

# NOTE: only for testing, not the answer of Q9
```{sql connection=dbcon}
SELECT MAX(total) 
  FROM (SELECT COUNT(t.trackId) AS total FROM tracks AS t
          INNER JOIN albums AS alb ON t.AlbumId = alb.AlbumId
          GROUP BY alb.ArtistId);
```

# Q10.Which artists have composed tracks in more than three different genres? 
# List the artists name and the number of distinct genres they have compositions in.
```{sql connection=dbcon}
SELECT art.Name, COUNT(DISTINCT t.GenreId) AS total FROM tracks AS t
  INNER JOIN albums AS alb ON t.AlbumId = alb.AlbumId
  INNER JOIN artists AS art ON alb.ArtistId = art.ArtistId
  GROUP BY art.ArtistId
  HAVING total > 3;
```

# NOTE: only for testing, not the answer of Q10
```{sql connection=dbcon}
SELECT DISTINCT art.Name, COUNT(DISTINCT t.GenreId) FROM tracks AS t
  INNER JOIN albums AS alb ON t.AlbumId = alb.AlbumId
  INNER JOIN artists AS art ON alb.ArtistId = art.ArtistId
  GROUP BY art.ArtistId;
```

```{r}
dbDisconnect(dbcon)
```
