---
title: "Query Databases and Dataframes in R with SQL"
output: html_notebook
### Author: Zhang, Jinwei
### Course: CS5200
### Term: 2023 Spring
---

# initialization
```{r}
library(RSQLite)
library(sqldf)
# get current work directory
wd = getwd()
fpath = paste0(wd, "/")
dbfile = "MediaDB.db"

# create db or get db connection
dbcon <- dbConnect(RSQLite::SQLite(), paste0(fpath, dbfile))
```

# Q1: insert column ExtPrice into invoice_items table, get the avg of inserted column
```{r}
sqlQuery <- 'SELECT * FROM invoice_items'
rs <- dbGetQuery(dbcon,sqlQuery)
rs$ExtPrice <- rs$Quantity * rs$UnitPrice
avg <- round(mean(rs$ExtPrice), 2)
rs
avg
```

# Q2: use sqldf to write query against rs
# display invoice ID, the total, and the number of items.
```{r}
rs2 <- sqldf('SELECT InvoiceId, SUM(ExtPrice) AS total, COUNT(*) AS num 
                FROM rs 
                GROUP BY InvoiceId')
rs2
```

# Q3: create a scatter plot
# total number of items in an invoice (x axis) versus the total (y axis)
```{r}
xData <- rs2$num
yData <- rs2$total
# change point to shape 19, remove frame of plot
plot(xData, yData, main = 'Invoice Plot',
     xlab = 'Number of Items', ylab = 'Total Amount',
     pch = 19, frame = FALSE)
```


# NOTE: only for testing -- drop the column first if exists
```{r}
alterStatement <- 'ALTER TABLE invoices DROP COLUMN DiscPrice'
dbExecute(dbcon, alterStatement)
```

# Q4
# add column DiscPrice to invoices table
```{r}
alterStatement <- 'ALTER TABLE invoices ADD DiscPrice NUMERIC'
dbExecute(dbcon, alterStatement)
```

# update records
# applies a 10% discount to the total amount for each invoice if it has more than 5 items
```{r}
updateStatement <- '
UPDATE invoices 
SET DiscPrice = Total * 0.9
WHERE InvoiceId IN(SELECT InvoiceId FROM invoice_items 
                   GROUP BY InvoiceId 
                   HAVING COUNT(*) > 5)
'
dbExecute(dbcon, updateStatement)

```

# Q5: show that the update in (4) executed properly by displaying a part of the table. 
```{r}
sqlQuery2 <- '
SELECT InvoiceId, CustomerId, Total, DiscPrice FROM invoices 
  WHERE DiscPrice IS NOT NULL 
  LIMIT 30'
rs4 <- dbGetQuery(dbcon,sqlQuery2)
rs4
```

```{r}
dbDisconnect(dbcon)
```
