---
title: "Build Triggers"
output: html_notebook
### Author: Zhang, Jinwei
### Course: CS5200
### Term: 2023 Spring
---

# create connection to db
```{r}
library(RSQLite)

# get current work directory
wd = getwd()
fpath = paste0(wd, "/")
dbfile = "asgmtDB.sqlite"

# create db or get db connection
dbcon <- dbConnect(RSQLite::SQLite(), paste0(fpath, dbfile))
```

# enable foreign key constraint
```{sql connection=dbcon}
PRAGMA foreign_keys = ON
```

# initialization
```{sql connection=dbcon}
DROP TABLE IF EXISTS Module;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS Affiliation;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS Author;
```

```{sql connection=dbcon}
DROP TABLE IF EXISTS AuthorToModule;
```

# create tables
```{sql connection=dbcon}
CREATE TABLE Module(
  number TEXT NOT NULL PRIMARY KEY,
  title TEXT NOT NULL,
  lengthInMin INTEGER NOT NULL
);
```

# TRIGGER 1
This trigger is for checking the value of attribute lengthInMin before inserting data into table Module 

If a module is created, the length of it must be positive number, any value smaller than 1 will be invalid.

Therefore, I create this trigger and use BEFORE INSERT to check the attribute value and abort the insert operation if the value is non-positive.
```{sql connection=dbcon}
CREATE TRIGGER IF NOT EXISTS check_lengthInMin_before_insert
  BEFORE INSERT ON Module
  BEGIN
    SELECT
      CASE
        WHEN NEW.lengthInMin < 1 THEN
          RAISE(ABORT, 'Invalid module length')
      END;
  END;
```

# Look-up table for categorical attribute in Author table
```{sql connection=dbcon}
CREATE TABLE Affiliation(
  aff_id INTEGER NOT NULL PRIMARY KEY AUTOINCREMENT,
  status TEXT NOT NULL UNIQUE
);
```

# insert predefined data
```{sql connection=dbcon}
INSERT INTO Affiliation(status) VALUES
  ("PT"),
  ("FT"),
  ("Contract");
```

```{sql connection=dbcon}
CREATE TABLE Author(
  aid INTEGER NOT NULL PRIMARY KEY AUTOINCREMENT,
  name TEXT NOT NULL,
  email TEXT NOT NULL,
  aff_id INTEGER NOT NULL,
  FOREIGN KEY (aff_id) REFERENCES Affiliation(aff_id)
);
```

# TRIGGER 2
This trigger is for checking the the value of attribute email before inserting data into table Author 

If an author is created, his/her email must follow the pattern '%_@_%._%'. In SQLite, '%_@_%._%' means there must be at least 1 character before '@', '.' and after '.'

In this way, the trigger can abort many inserting operations with email being obviously wrong, which reduces the logic complexity of application server.

Therefore, I create this trigger and use BEFORE INSERT to check the email value and abort the inserting operation if the value doesn't follow the valid email format
```{sql connection=dbcon}
CREATE TRIGGER IF NOT EXISTS check_email_before_insert
  BEFORE INSERT ON Author
  BEGIN
    SELECT
      CASE
        WHEN NEW.email NOT LIKE '%_@_%._%' THEN
          RAISE(ABORT, 'Invalid email format')
      END;
  END;
```

# TRIGGER 3
Similar to TRIGGER 2, we should also check the email before updating data in table Author
```{sql connection=dbcon}
CREATE TRIGGER IF NOT EXISTS check_email_before_update
  BEFORE UPDATE ON Author
  BEGIN
    SELECT
      CASE
        WHEN NEW.email NOT LIKE '%_@_%._%' THEN
          RAISE(ABORT, 'Invalid email format')
      END;
  END;
```

# mapping table for Author and Module
```{sql connection=dbcon}
CREATE TABLE AuthorToModule(
  a2m_id INTEGER NOT NULL PRIMARY KEY AUTOINCREMENT,
  number TEXT NOT NULL,
  aid INTEGER NOT NULL,
  FOREIGN KEY (number) REFERENCES Module(number),
  FOREIGN KEY (aid) REFERENCES Author(aid)
);
```

# insert data for testing
```{sql connection=dbcon}
INSERT INTO Module(number, title, lengthInMIN) VALUES
  ("1", "module1", 30),
  ("2", "module2", 45),
  ("3", "module3", 50);
```

# test trigger 1: invalid lengthInMin value
```{sql connection=dbcon}
INSERT INTO Module(number, title, lengthInMIN) VALUES
  ("4", "module4", -10);
```

```{sql connection=dbcon}
INSERT INTO Author(name, email, aff_id) VALUES
  ("Mike", "mike@test.com", 1),
  ("Siri", "siri@test.com", 1),
  ("Jerry", "jerry@test.com", 2),
  ("Tom", "tom@test.com", 3);
```

# test trigger 2: insert record with invalid email value
```{sql connection=dbcon}
INSERT INTO Author(name, email, aff_id) VALUES
  ("Kyrie", "kyrie.com", 1);
```

# test trigger 3: update email with invalid value
```{sql connection=dbcon}
UPDATE Author 
  SET email = 'mike@gmail'
  WHERE name = 'Mike';
```

```{sql connection=dbcon}
INSERT INTO AuthorToModule(number, aid) VALUES
  ("1", 1),
  ("2", 2),
  ("3", 3);
```

```{sql connection=dbcon}
SELECT * from Module;
```

```{sql connection=dbcon}
SELECT * from Affiliation;
```

```{sql connection=dbcon}
SELECT * FROM Author;
```

```{sql connection=dbcon}
SELECT * FROM AuthorToModule;
```

# test join
```{sql connection=dbcon}
SELECT Author.aid, Author.name, Author.email, Affiliation.status, Module.title, Module.lengthInMin FROM AuthorToModule
  INNER JOIN Author ON Author.aid = AuthorToModule.aid,
             Module ON Module.number = AuthorToModule.number,
             Affiliation ON Author.aff_id = Affiliation.aff_id;
```

# discconect db
```{r}
dbDisconnect(dbcon)
```

